Model exploration

### 2015-03-12 documentation/model_exploration.Rmd
update model exploration

---
title: "Model exploration"
author: "Florian D. Schneider"
date: "Tuesday, February 25th, 2014"
output:
  pdf_document:
    fig_height: 3
    fig_width: 7
fontsize: 12pt
geometry: margin=1.4in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

# alternative stable states due to feedbacks

@@

## local mechanisms

@@

### facilitation

The mechanism of facilitation, i.e. the positive effect of one individual or species onto another, is well described in terms of species interaction networks [@bruno09; @kefi12]. Such mechanisms can be species-pair specific but they also play a general abiotic role as the cause of local amelioration of the environmental conditions of the neighborhood of a 'nurse' plant.

@@

bifurcation(parms, pairapprox = TRUE, "b", c(0,1))

@@

### local competition

@@

parms$c = 0.3

@@

bifurcation(parms, pairapprox = TRUE, "b", c(0,1))

@@

### associational resistance

@@

parms$L = 8
parms$p = 0.7

@@

bifurcation(parms, pairapprox = TRUE, "L", c(0,12))

@@

### attractant-decoy

@@

@@

@@

We conclude that the different kinds of local feedbacks can fundamentally alter the critical thresholds of arid rangelands. Knowing about their relative importance is vital for the management of arid ecosystems under grazing pressure.

>>>>

---
title: "Model exploration"
author: "Florian D. Schneider"
date: "Tuesday, February 25th, 2014"
output:
  pdf_document:
    fig_height: 3
    fig_width: 7
    fig_caption: yes
    fig_crop: no
    includes:
      in_header: include.tex
fontsize: 12pt
geometry: margin=1.4in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

## alternative stable states due to global feedbacks

@@

## local feedback mechanisms

@@

#### facilitation

The mechanism of facilitation, i.e. the positive effect of one individual or species onto another, is well described in terms of species interaction networks [@bruno09; @kefi12]. Such mechanisms can be species-pair specific but they also play a general abiotic role as the cause of local amelioration of the environmental conditions of the neighborhood of a 'nurse' plant. The mechanisms commonly subsumised under local facilitation are shading effects, soil water retention, aggregation of litter and fixation of organic matter in top soils [@bautista].

We implement local facilitation as a mechanism that increases growth rates on empty ground locally with increasing local plant cover, $q_{0|1}$.

```{r, echo = FALSE, fig.cap="a) mean field approximation of density dependent mortality and growth of vegetation with local facilitation. As plant cover increases, the effect of facilitation is stronger. solid points show steady states, open circle shows unstable equilibrium. b) bifurcation diagram of vegetation cover over a gradient of environmental quality (black: steady states, grey: unstable equilibria). High facilitation has strong stabilising effects on the vegetated state. "}

@@

parms$f = 0.8
bifurcation(parms, pairapprox = FALSE, "b", c(0,1))

parms$f = 0.4
bifurcation(parms, pairapprox = FALSE, "b", c(0,1), add =TRUE)

parms$f = 0
bifurcation(parms, pairapprox = FALSE, "b", c(0,1), add =TRUE)

#parms$b = 0.5
#parms$f = 0.8
#bifurcation(parms, pairapprox = FALSE, "L", c(0,12))
#
#parms$f = 0.4
#bifurcation(parms, pairapprox = FALSE, "L", c(0,12), add = TRUE)
#
#parms$f = 0
#bifurcation(parms, pairapprox = FALSE, "L", c(0,12), add = TRUE)

@@

Facilitation has a strong stabilizing effect on the vegetated state of the ecosystem and extends the range where vegetation can persist far towards very harsh environmental conditions.

#### local competition

Local competition is a mechanism counteracting local facilitation. If plants grow close together they exploit the locally available resources, like water or nutrients. Thus, the perceived carrying capacity of an empty space is reduced as the neighboring plant cover $q_{1|1}$ increases.

@@

parms$c = 0.6

@@

parms$c = 0
bifurcation(parms, pairapprox = FALSE, "b", c(0,1))

parms$c = 0.3
bifurcation(parms, pairapprox = FALSE, "b", c(0,1), add = TRUE)

parms$c = 0.6
bifurcation(parms, pairapprox = FALSE, "b", c(0,1), add = TRUE)

@@

Where local competition is important, the total vegetation cover is reduced, since competition affects growth primarily at high cover. The critical thresholds of sustaineable environmental quality are hardly affected.

#### associational resistance

In arid rangelands many plants develop protective physical structures, like thorns or cushion growth, or secondary plant metabolites to reduce the impact of grazing. This investment usually not only benefits the plant itself but also other plants in the direct neighborhood, since it reduces the accessibility of space or acts as a repellent for herbivores. 
This associational resistance is providing a local decrease of search efficiency (or an increase of handling effort) for grazers, which is reducing total consumption, and thus acts as a positive local feedback on vegetation cover.

@@

parms$L = 6
parms$p = 0.6

@@

parms$p = 0.0
bifurcation(parms, pairapprox = FALSE, "L", c(0,12))
text(3,0.7, "0.0")

parms$p = 0.6
bifurcation(parms, pairapprox = FALSE,add = TRUE, "L", c(0,12))
text(7,0.65, "0.6")

parms$p = 1
bifurcation(parms, pairapprox = FALSE,add = TRUE, "L", c(0,12))
text(8,0.81, "1.0")

#parms$p = 0.0
#bifurcation(parms, pairapprox = FALSE, "b", c(0,1))
#text(3,0.7, "0.0")

#parms$p = 0.6
#bifurcation(parms, pairapprox = FALSE,add = TRUE, "b", c(0,1))
#text(7,0.65, "0.6")

#parms$p = 1
#bifurcation(parms, pairapprox = FALSE,add = TRUE, "b", c(0,1))
#text(8,0.81, "1.0")

@@

Associational resistance is a stabilizing feature of vegetation cover. It greatly extends the range of persistent vegetation cover under high levels of livestock grazing, since it effectively reduces grazing impact at high vegetation cover.

#### attractant-decoy

As an opposing mechanism to associational resistance, palatable plants attract grazers and

@@

parms$v = 0.9

@@

text(8,0.75, "0.9")

parms$v = 0.5

bifurcation(parms, pairapprox = FALSE, "L", c(0,12), add = TRUE)

text(8,0.75, "0.5")

parms$v = 0

bifurcation(parms, pairapprox = FALSE, "L", c(0,12), add = TRUE)

text(8,0.75, "0")

@@

Note that competition too could cause increased mortality due to high local cover. For reasons of simplicity we do not mix those effects here, but consider it of similar effect as an inverse of the associational resistance.

### conclusion

We conclude that the different kinds of local feedbacks can fundamentally alter the stability properties of arid rangelands. Knowing about their relative importance is vital for the management of arid ecosystems under grazing pressure.
